---
title: "Introduction to BayesianReasoning"
author: "Gorka Navarrete"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BayesianReasoning}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
params:
  package_creation: FALSE
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 60,
  fig.height = 10,
  fig.width = 14
)
```

```{r, echo = FALSE, message = FALSE, results = 'hide'}
library(BayesianReasoning)

# FROM: https://community.rstudio.com/t/internet-resources-should-fail-gracefully/49199/12
safely_show_image_Rmd <- function(remote_file, name_image) {
  try_GET <- function(x, ...) {
    tryCatch(
      httr::GET(url = x, httr::timeout(1), ...),
      error = function(e) conditionMessage(e),
      warning = function(w) conditionMessage(w)
    )
  }
  is_response <- function(x) {
    class(x) == "response"
  }
  
  # First check internet connection
  if (!curl::has_internet()) {
    message("No internet connection.")
    return(invisible(NULL))
  }
  # Then try for timeout problems
  resp <- try_GET(remote_file)
  if (!is_response(resp)) {
    message(resp)
    return(invisible(NULL))
  }
  # Then stop if status > 400
  if (httr::http_error(resp)) { 
    httr::message_for_status(resp)
    return(invisible(NULL))
  }
  
  # Output
  paste0("![", name_image, "](", remote_file, ")")
}

```

# Bayesian reasoning

 
<!-- badges: start -->

```{r echo=FALSE, results='asis'}

# If we are creating the README params$package_creation will be TRUE 
if (params$package_creation) {

  cat(paste0("[", safely_show_image_Rmd(remote_file = "https://www.r-pkg.org/badges/version/BayesianReasoning", name_image = "CRAN status"), "]", "(https://cran.r-project.org/package=BayesianReasoning)"))
  cat(paste0("[", safely_show_image_Rmd(remote_file = "https://codecov.io/gh/gorkang/BayesianReasoning/branch/master/graph/badge.svg", name_image = "Codecov test coverage"), "]", "(https://app.codecov.io/gh/gorkang/BayesianReasoning?branch=master)"))
  cat(paste0("[", safely_show_image_Rmd(remote_file = "http://cranlogs.r-pkg.org/badges/BayesianReasoning", name_image = "downloads"), "]", "(https://cran.r-project.org/package=BayesianReasoning)"))
  cat(paste0("[", safely_show_image_Rmd(remote_file = "https://img.shields.io/badge/lifecycle-experimental-orange.svg", name_image = "Lifecycle: experimental"), "]", "(https://lifecycle.r-lib.org/articles/stages.html#experimental)"))
  cat(paste0("[", safely_show_image_Rmd(remote_file = "https://zenodo.org/badge/93097662.svg", name_image = "DOI"), "]", "(https://zenodo.org/badge/latestdoi/93097662)"))

}
```

<!-- badges: end -->

---  

## Bayesian reasoning in medical contexts


This package includes a few functions to plot and help understand Positive and Negative Predictive Values, and their relationship with Sensitivity, Specificity and Prevalence.  

+ The **Positive Predictive** Value of a medical test is the probability that a positive result will mean having the disease. Formally p(Disease|+)
+ The **Negative Predictive** Value of a medical test is the probability that a negative result will mean **not** having the disease. Formally p(Healthy|-)  


The BayesianReasoning package has three main functions:  

+ **PPV_heatmap()**: Plot heatmaps with PPV or NPV values for the given test and disease parameters.
+ **PPV_diagnostic_vs_screening()**: Plots the difference between the PPV of a test in a diagnostic context (very high prevalence; or a common study sample, e.g. ~50% prevalence) versus a screening context (lower prevalence).
+ **min_possible_prevalence()**: Calculates how high should the prevalence of a disease be to reach a desired PPV given certain test parameters.  


--- 

If you want to install the package can use: `remotes::install_github("gorkang/BayesianReasoning")`. Please report any problems you find in the [Issues Github page](https://github.com/gorkang/BayesianReasoning/issues).  

There is a [shiny app implementation](https://gorkang.shinyapps.io/BayesianReasoning/) with most of the main features available.  

---


## PPV_heatmap()

Plot heatmaps with PPV or NPV values for a given specificity and a range of Prevalences and FP or FN (1 - Sensitivity). The basic parameters are:

* min_Prevalence: Min prevalence in y axis. "min_Prevalence out of y"  
* max_Prevalence: Max prevalence in y axis. "1 out of max_Prevalence"  
* Sensitivity: Sensitivity of the test  
* max_FP: FP is 1 - specificity. The x axis will go from FP = 0% to max_FP  
* Language: "es" for Spanish or "en" for English  

```{r heatmap}

PPV_heatmap(min_Prevalence = 1,
            max_Prevalence = 1000, 
            Sensitivity = 100, limits_Specificity = c(90, 100),
            Language = "en")
            
```  




---   

### NPV

You can also plot an NPV heatmap with PPV_NPV = "NPV".  

```{r NPV-heatmap}

PPV_heatmap(PPV_NPV = "NPV",
            min_Prevalence = 800, max_Prevalence = 1000, 
            Specificity = 95, limits_Sensitivity = c(90, 100),
            Language = "en")

```



--- 


### Area overlay

You can add different types of overlay to the plots. 

For example, an area overlay showing the point PPV for a given prevalence and FP or FN:  


```{r area}

PPV_heatmap(min_Prevalence = 1, max_Prevalence = 1200, 
            Sensitivity = 81, 
            limits_Specificity = c(94, 100),
            label_subtitle = "Prenatal screening for Down Syndrome by Age",
            overlay = "area",
            overlay_labels = "40 y.o.",
            overlay_position_FP = 4.8,
            overlay_prevalence_1 = 1,
            overlay_prevalence_2 = 68)

```

The area plot overlay can show more details about how the calculation of PPV/NPV is performed:  

```{r area2}

PPV_heatmap(min_Prevalence = 1, max_Prevalence = 1200, 
            Sensitivity = 81, 
            limits_Specificity = c(94, 100),
            label_subtitle = "Prenatal screening for Down Syndrome by Age", 
            overlay_extra_info = TRUE,
            overlay = "area",
            overlay_labels = "40 y.o.",
            overlay_position_FP = 4.8,
            overlay_prevalence_1 = 1,
            overlay_prevalence_2 = 68)

```




### Line overlay

Also, you can add a line overlay highlighting a range of prevalences and FP. This is useful, for example, to show how the PPV of a test changes with age:  


```{r line}

PPV_heatmap(min_Prevalence = 1, max_Prevalence = 1800, 
            Sensitivity = 90, 
            limits_Specificity = c(84, 100),
            label_subtitle = "PPV of Mammogram for Breast Cancer by Age",
            overlay = "line", 
            overlay_labels = c("80 y.o.", "70 y.o.", "60 y.o.", "50 y.o.", "40 y.o.", "30 y.o.", "20  y.o."),
            overlay_position_FP = c(6.5, 7, 8, 9, 12, 14, 14),
            overlay_prevalence_1 = c(1, 1, 1, 1, 1, 1, 1),
            overlay_prevalence_2 = c(22, 26, 29, 44, 69, 227, 1667))
                
```


---   

Another example. In this case, the FP is constant across age:

```{r line-2}

PPV_heatmap(min_Prevalence = 1, max_Prevalence = 2000, Sensitivity = 81, 
            limits_Specificity = c(94, 100),
            label_subtitle = "Prenatal screening for Down Syndrome by Age",
            overlay = "line",
            overlay_labels = c("40 y.o.", "30 y.o.", "20 y.o."),
            overlay_position_FP = c(4.8, 4.8, 4.8),
            overlay_prevalence_1 = c(1, 1, 1),
            overlay_prevalence_2 = c(68, 626, 1068))
                
```


---   

## PPV_diagnostic_vs_screening()

In scientific studies developing a new test for the early detection of a medical condition, it is quite common to use a sample where 50% of participants has a medical condition and the other 50% are normal controls. This has the unintended effect of maximizing the PPV of the test.  

This function shows a plot with the difference between the PPV of a diagnostic context (very high prevalence; or a common study sample, e.g. ~50% prevalence) versus that of a screening context (lower prevalence).  


```{r diagnostic}

PPV_diagnostic_vs_screening(max_FP = 10, 
                            Sensitivity = 100, 
                            prevalence_screening_group = 1000, 
                            prevalence_diagnostic_group = 2)
                            
```  




---   

## min_possible_prevalence()

Imagine you would like to use a test in a population and want to have a 98% PPV. That is, IF a positive result comes out in the test, you would like a 98% certainty that it is a true positive.  

How high should the prevalence of the disease be in that group?  

```r 
min_possible_prevalence(Sensitivity = 100, 
                        FP_test = 0.1, 
                        min_PPV_desired = 98)
```

> To reach a PPV of 98 when using a test with 100 % Sensitivity and 0.1 % False Positive Rate, you need a prevalence of at least 1 out of 21

--- 

Another example, with a very good test, and lower expectations:  

```r 
min_possible_prevalence(Sensitivity = 99.9, 
                        FP_test = .1, 
                        min_PPV_desired = 70)
```

> To reach a PPV of 70 when using a test with 99.9 % Sensitivity and 0.1 % False Positive Rate, you need a prevalence of at least 1 out of 429

--- 

